pdp: An R Package for Constructing Partial Dependence Plots

### 2016-12-13 manuscript/reviews/response.Rmd

---
title: "Response to Comments"
output:
  html_document: default
  html_notebook: default
  pdf_document: default
  word_document: default
---

Thanks to the reviewers and editor for the well thought out comments. I believe that they have greatly imporved, not only the manuscript, but the package it describes as well. I hope that my responses are at least satisfactory. Please find below a copy of the reviewers comments along with my responses.

@@

#### Comment
The following code gives an obscure error message:
```
"Error in seq.default(from = min(train[[x]], na.rm = TRUE),
   to = max(train[[x]],  :
  'from' cannot be NA, NaN or infinite"
```
Some better handholding for the user may be helpful here.
```{r, eval=FALSE}
library(pdp)
data(trees)
mod <- lm(Volume~., data=trees)
partial(mod, pred.var="nosuchvariable", plot=TRUE) # obscure err msg
```

#### Response
This has been fixed in `pdp` (version 0.3.0) by adding an informative error message listing which variables are the cause of the issue.
```{r, eval=FALSE}
library(pdp)
data(trees)
mod <- lm(Volume~., data=trees)
partial(mod, pred.var=c("Girth", "foo", "bar"), plot=TRUE)
```
```
Error in partial.default(mod, pred.var = "nosuchvariable", plot = TRUE) : 
  nosuchvariable not found in the training data.
```

#### Comment
On page 2 of the paper it says "The columns of the data frame are labeled in the same order as the features supplied to `pred.var`, and the last column is always labeled `y`". What happens if `y` is one of the predictors?

#### Response
The last column is labeled `y` for consistency with `gbm` and `randomForest` behavior. Unfortunately, both `pdp::partial` and `randomForest::partialPlot` fail in this case; `gbm`, however, does not fail. I think regardless of what the default column name is, there is always the possibility of this issue. I filed an issue on the GitHub issue page: https://github.com/bgreenwell/pdp/issues/20. In the mean time, `partial` will throw an error describing the problem so that the user can take action.

@@

#### Response
Fixed.

#### Comment
On page 3 in the PDF of the paper, add a newline to the error message in the paper:
```
Error: The training data could not be extracted from object. Please
supply the raw training data using the ...
```

#### Response
Fixed.

#### Comment
Is there any reason the package is called `pdp` but the main function is called `partial`?  Consider giving the function the same name as the package, so e.g., the following works (to help the user get started)
```{r, eval=FALSE}
library(pdp)
?pdp # fails
```

#### Response
The package was originally called `partial`, but at the time of its original submission this conflicted with a CRAN package of the same name (but with mixed lowercase/uppercase letetrs). So, I decided to change the name of the package to `pdp`; besides, I think `partial` and `plotPartial` work well together as function names. However, I did add a help page describing the package, main functions, etc. So, `?pdp` no longer fails. I am, however, not opposed to creating an alias to `partial` called `pdp` in the future.

@@

#### Comment
`partial` should return the same value regardless of whether or not the `plot` argument is used?

#### Response
I dont think so and this is the same behavior for the PDP functions in `randomForest` and `gbm` (not that `pdp` needs to immitate the behavior of these packages PDP functions). However, I did add the partial dependence data as an attribute to the returned `"trellis"` object (i.e., plot) so that the user can extract the data easily from that if necessary (e.g., if they do not like the plot and don't want to re-run partial). For example,
```{r}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
pd.Girth <- partial(mod, pred.var = "Girth")  # data only
pdp.Girth <- partial(mod, pred.var = "Girth", plot = TRUE)  # plot
identical(pd.Girth, attr(pdp.Girth, "partial.data"))  # should be identical
```
Also the `lattice` function `trellis.last.object` is now exported when `pdp` is loaded. This will be useful when the user does not store the resulting plot in the call to `partial` but needs to retrieve the outout data to avoid re-running `partial` when computation time is a concern. For example,
```{r}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
partial(mod, pred.var = "Girth", plot = TRUE)
pdp.Girth <- lattice::trellis.last.object()
pd.Girth <- attr(pdp.Girth, "partial.data")
head(pd.Girth)
```

@@

#### Comment
A call to say `par(mfrow = c(2, 2))` before calling `plotPartial` is ignored by `plotPartial`.  This is a pity because it would allow multiple plots to be put on one page like this:
```{r, eval=FALSE}
par(mfrow=c(2,2))
partial(ozone.rf, pred.var = "Temp", plot = TRUE)
partial(ozone.rf, pred.var = "Wind", plot = TRUE)
```

#### Response
The command `par(mfrow = c(2, 2))` does not work with `"trellice"` objects, such as those produced by `lattice` which `partial` relies on for producing its plots. Other methods are available. For example, using the `gridExtra` package:
```{r, eval=TRUE}
library(randomForest)
data(airquality)
ozone.rf <- randomForest(Ozone ~ ., data=airquality, na.action=na.omit)
library(pdp)
pdp1 <- partial(ozone.rf, pred.var = "Temp", plot = TRUE)
pdp2 <- partial(ozone.rf, pred.var = "Wind", plot = TRUE)
gridExtra::grid.arrange(pdp1, pdp2, ncol = 2)
```
For convenience, `pdp` now imports the `grid.extra` function and this has been incorporated into the manuscript where appropriate.

@@

#### Comment
It would be nice if the functions automatically plotted all the (important) variables in a grid of plots, without forcing the user to explicitly specify them.

#### Response 
I agree, this would be a nice feature, but the problem I see is that not all models emit a "variable importance". The `caret` package has a come a loong way in this regard. At this point, I think it is simpler to have the user select which variables they would like to see plotted (it forces them to explore the results). I am not opposed to including this feature in the future. I added this as a feature request: https://github.com/bgreenwell/pdp/issues/19.

@@

#### Comment
It would be nice if the functions automatically got the data from the model used to build the model in more cases without forcing the user to explicitly specify the data.  Since XGBoost seems now quite popular, perhaps the authors of that package could be persuaded to make the original model data available for functions like partial (currently XGBoost uses a custom object `xgb.DMatrix` which is currently write-only -- one can't access the data stored in `xgb.DMatrix`).

#### Response
This is also true for the fitting functions in other package (e.g., `kernlab::ksvm`). As time goes on, the coverage will certainly be inncreased to make it easier on the user. For example, `partial` now works with `caret` which always stores the taining data in a component called `trainingData`. I suspect that the package maintainers for `xgboost`, would be hesitent in storing a copy of the training data, especially since the package is often used for modeling large data sets; they probably do not want to store a copy of the data with the fitted model object for memory reasons.

@@

#### Comment
Consider adding a `slowtests` directory (below `pdp/inst/` or elsewhere) that has tests for all the models supported by the package (Table 1 in the paper) to easily check back compatibility when packages change.  (For justification for such tests see for example my comments on `gbm` back compatibility above.)  These tests will probably be quite slow, but that won't be an issue because they won't be invoked by CRAN check; the slow tests would be just for the maintainer of the `pdp` package to run manually.

#### Response
This is a good suggestion and is currently under development. In fact, the `gbm` maintainers have already contacted me about a breaking change caused by the next planned release of their package.

@@

## Reviewer 2

#### Comment 
The Introduction immediately introduces the PDP. You must first motivate why it is important. In a linear model, they are not important. In a linear model with simple interactions, likewise. They first become important if you have non-parametric machine-learning techniques. Further, they are not the only idea extant when trying to explain a predictor's impact on the response. This paper would benefit if you can give a bit more of a lit review here.

#### Response
I agree that PDPs are not as useful in "simple" linear models (e.g., few variables, only two-way interactions, etc.), but certainly there will be cases where PDPs will be extremely useful (e.g., many terms, complex interactions, etc.). In fact, linear models are still common in machine learning/computer science/data science (they just typically contain many terms). I also agree that PDPs are not the only, or sometimes best, approach to understanding the impact of a subset of the predictor's on the response. Brief descriptions and references are given for many alternatives: marginal model plots (`plotmo`), individual conditional expectation plots (`ICEbox`)---which are described as an alternatve to PDPs when strong interactions are suspected, effect plots (`effects`), and a few others available in the `car` package. Unless the editor disagrees, I don't think it is necessary to go into more detail than what is already provided at the end of the introduction.

@@

### Important Clarifications Required:

#### Comment
Top of p2. The `ICEbox` package allows for general construction and visualization of PDP's. However, it is limited as you only get one dimensional representations. The selling point of your package should focus on the flexibility of graphics and the illustration of non-extrapoloation zones.

#### Response
ICEbox does not actually construct PDPs. Only graphics trying to do the same thing: visualize the relationship between a predictor and the response. I think flexible graphics and non-extrapoloation zones are only half of the selling point here. I think that being able to run `partial` in parallel is huge, and also that this packge works with many types of machine learning models---not just tree-based models. For example, as described in the manuscript, `pdp` can produce single or multi-predictor PDPs for conditional random forests and XGBoost models (among many others), and in parallel. I expanded the last example to describe the interface with `caret`, which I think opens the door for wonderful exploratory data analysis pipe lines (e.g., choose model --> plot importance --> plot relationships, etc.). Most, if any, of these other software packages do not describe or support these important components (e.g., parallel execution), not even ICEbox.

@@

#### Comment
All `pdp` figures throughout the paper: Can you make explicit in the caption what code produces the left and right subfigures? I would use the subfigure package in latex and then put a comment in the R code e.g. # Fig 3b.

#### Response
Rather than using subfigures, which I think would look messy with all the code, I made quite an attempt to comment which pieces of code produce which figure---in addition to in the text. Also, in response to another reviewer's comments, the code which produces figures containing multiple plots is better described and demonstrated. See, for example, the section called Multi-predictor PDPs.

@@

### Edits:

#### Comment
p1 abstract: strict assumptions imposed by traditional statistical models

#### Comment
p1 pairs should be using curly braces of angle braces, not square braces as this is easily confused with closed intervals.
#### Response
Fixed. Switched to curly braces.

#### Comment
p2 unless the editor disagrees, there is no reason to add "install.packages("pdp")" in the text of the paper as it is assumed a user knows how to install a package.
#### Response
Fixed. Code chunk was removed.

#### Comment
p3 "Error: The training..." should be wrapped.
#### Response
Fixed. The text is now wrapped appropriately.

#### Comment
Fig 6: the sidebar should be labeled "ozone". Also, would it be terrible to label all axes even though it is redundant?
#### Response
The sidebar (i.e., colorkey) has been labeled, but for this example only. As sad as it sounds, labeling the colorkey is not as trivial as we'd expect for lattice plots. The code for doing so is incorporated in the example in question. Also, the text describing the first code chunck that produces a display with a colorkey now contains a footnote pointing to the example where the code to produce a label can be found. This is also motivation for why one of the examples showed how to use `ggplot2` for plotting the results from `partial`---`ggplot2` makes it a trivial matter to label the colorkey.

@@

#### Comment
Fig 7: as beautiful as this is, you should label the sidebar as "logit" or "centered logit" instead of "y".
#### Response
Fixed. Color key labeled "Average logit".

### Suggestions for the software

#### Comment
I understand that you like the feature of having a one-shot function partial to both generate and plot the PDP. I realize that the default is `FALS`E. However, I still think it's a disservice to the user as constructing PDP's is computationally intense and if they get the plot wrong or it's not to their liking, they will have to regenerate the whole thing again. So, it is my opinion that you should take it out and force two steps: generation and illustration.
#### Response
While I am not suggesting that the `partial` function needs to mimic the behavior of the PDP fucntions in `randomForest` and `gbm`, I do find it useful to be able to plot the results without forcing the user to store them. However, I agree with your concern which is why `partial` has `plot = FALSE` as the default. Furthermore, I have made it possible to retrieve the raw PDP data even when \code{plot = TRUE}. This is because the "trellis" object returned by `partial` \ `plotPartial` now contains an additional `partial.data` attribute. This is described in the package documentation. For example,
```{r, eval=TRUE}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
partial(mod, pred.var = "Girth", plot = TRUE)
pdp.Girth <- lattice::trellis.last.object()
pd.Girth <- attr(pdp.Girth, "partial.data")
head(pd.Girth)
```

#### Comment
The package could be more powerful if you allowed a predict function as a parameter. For instance, you cannot currently use random forests to illustrate classification probabilities.
#### Response
I agree, and I have changed the behavior of `...` to refer to additional parameters to `stats::predict` (e.g., the `n.trees` argument in `predict.gbm`.)

#### Comment
p7 I believe there are packages allowing for parallelization in Windows. Can you switch the example so this works on all platforms?
#### Response
Good suggestion. I tweaked the example to run in both Windows and Unix-like systems. This is also reflected in the manuscript text.

### 2016-12-13 manuscript/greenwell.tex
tweaks based on review comments

\abstract{
Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes $E\left[\boldsymbol{\mathcal{Y}}\right] = \boldsymbol{X}\boldsymbol{\beta}$). Unfortunately, it can be challenging to understand the results of such models and explain them to management. Partial dependence plots offer a simple solution. Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be
more easily understood. These plots are especially useful in explaining the output from black box models. In this paper, we introduce \pkg{pdp}, a general R package for constructing partial dependence plots.
}

@@

\strong{Note:} as mentioned above, \pkg{pdp} relies on \pkg{lattice} for its graphics. \pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}. \pkg{grid} graphics behave a little differently than traditional R graphics and two points are worth making (see \code{?lattice} for more details):
\begin{enumerate}
  \item \pkg{lattice} functions return a \code{"trellis"} object, but do not display it; the \code{print} method produces the actual display. However, due to R's automatic printing rule, the result is automatically printed when using these functions in the command line. If \code{plotPartial} is called inside of \code{source} or inside a loop (e.g., \code{for} or \code{while}), an explicit \code{print} statement is required to display the resulting graph.
  \item Setting graphical parameters via \code{par} typically has no effect on \pkg{lattice} plots. Instead, \pkg{lattice} provides its own \code{trellis.par.set} function for modifying graphical parameters.
\end{enumerate}
A consequence of second point is that the \code{par} function cannot be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots. Simple solutions are available in packages \CRANpkg{latticeExtra} and \CRANpkg{gridExtra}. For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{pdp} plots on a single page.

@@

When additional arguments are necessary, the user will be prompted with an informative error message. For instance, the user may occasionally see the following:
\begin{example}
Error: The training data could not be extracted from object. Please supply
the raw training data using the `train` argument in the call to `partial`.
\end{example}
This will occur, for example, when using partial with the \code{"xgb.Booster"} class. In fact, using \code{partial} with any object that does not store the original training data will result in the above error message.

The \code{partial} function also supports object of class \code{"train"} produced using the \code{train} function from the well-known \CRANpkg{caret} package \citep{caret-pkg}. This means that \code{partial} can be used with any classification or regression model that \code{caret} supports; see \url{http://topepo.github.io/caret/available-models.html} for a current list of supported models. An example is given in Section~?.

For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}. These data contain the median value of owner-occupied homes in 506 U.S. census tracts in the Boston area, along with 13 independent variables such as the per capita crime rate by town. We begin by loading the data and omitting unimportant columns.
\begin{example}
data(BostonHousing2, package = "mlbench")  # mlbench must be installed!
boston <- BostonHousing2[, -c(1, 2, 5)]
\end{example}
Next, we fit a random forest to the entire data set with default tuning parameters and 500 trees:
\begin{example}
library(randomForest)
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # variable importance plot
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}). The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk uses the random forest model to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}. The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
rwb <- colorRampPalette(c("red", "white", "blue"))
pd <- partial(boston.rf, pred.var = c("lstat", "rm"))
plotPartial(pd)  # Figure 3 (left)
plotPartial(pd, contour = TRUE, col.regions = rwb)  # Figure 3 (middle)
plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = TRUE, screen = list(z = -20, x = -60))  # Figure 3 (right)
\end{example}
\strong{Note:} the default color map for level plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Addressing computational concerns
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Addressing computational concerns}

Constructing PDPs can be quite computationally expensive. Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option, then uses \code{grid.arrange} to display all three PDPs on the same page; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
p1 <- partial(boston.rf, "rm", plot = TRUE)
p2 <- partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE)
p3 <- partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE)
grid.arrange(
  p1,  # Figure 5 (left)
  p2,  # Figure 5 (middle)
  p3,  # Figure 5 (right)
  ncol = 3
)
\end{example}

@@

Setting up a parallel backend is rather straightforward. To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on both Windows and Unix-like systems using the \pkg{doParallel} package.
\begin{example}
library(doParallel)  # load parallel backend
cl <- makeCluster(4)  # use 4 workers
registerDoParallel(cl)
\end{example}
Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} and \code{paropts} options and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded\footnote{Notice we have to pass the names of external packages that the tasks depend on via the \code{paropts} argument; in this case, \code{"earth"}. See \code{?plyr::adply} for details.}. This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel. The result is displayed in Figure~\ref{fig:partial_par}. \strong{Note:} it is considered good practice to shut down the workers by calling \code{stopCluster} when finished.
\begin{example}
partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
        chull = TRUE, parallel = TRUE, paropts = list(.packages = "earth"))  # Figure 6
stopCluster(cl)  # good practice
\end{example}

@@

To round out our discussion, we provide one last example using a recently popular (and successful!) machine learning tool. XGBoost, short for eXtreme Gradient Boosting, is a popular library providing optimized distributed gradient boosting that is specifically designed to be highly efficient, flexible and portable. The associated R package \pkg{xgboost} has been used to win a number of \href{https://www.kaggle.com/}{Kaggle competitions}. It has been shown to be many times faster than the well known \pkg{gbm} package. However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for constructing PDPs. Fortunately, the \pkg{pdp} package can be used to fill this gap.

For illustration, we return to the Boston housing data. The code chunk below fits an \pkg{xgboost} model to the \code{boston} data frame from Section~\ref{sec:boston}; the tuning parameters were predetermined using 5-fold cross-validation. (Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.)

The code chunk below uses \pkg{caret} to train an XGBoost model over a grid of tuning values using 5-fold cross-validation.
\begin{example}
# Optimize tuning parameters using 5-fold cross-validation
library(caret)

@@

# Grid of tuning values for XGBoost
xgb.grid <- expand.grid(nrounds = c(100, 500, 1000, 2000, 5000),
                        max_depth = 1:6,
                        eta = c(0.001, 0.01, 0.1, 0.5, 1),
                        gamma = 0,
                        colsample_bytree = 1,
                        min_child_weight = 1)

# Train an XGBoost model
set.seed(202)
boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
                    y = boston$cmedv,
                    method = "xgbTree",
                    metric = "RMSE",
                    trControl = ctrl,
                    tuneGrid = xgb.grid)
\end{example}
The model with the smallest root-mean-squared error (RMSE) had \code{nrounds = 2000}, \code{max\_depth = 3}, \code{eta = 0.01}, and achieved a cross-validated RMSE and $R^2$ of 2.819 and 0.905, respectively. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display al three PDPs on the same page. The results are displayed in Figure~\ref{fig:boston_xgb}.
\begin{example}
p1 <- partial(boston.xgb, pred.var = "rm", plot = T, rug = T)
p2 <- partial(boston.xgb, pred.var = "lstat", plot = T, rug = T)
p3 <- partial(boston.xgb, pred.var = c("lstat", "rm"), plot = T, chull = T),
grid.arrange(
  p1,  # Figure 8 (left)
  p2,  # Figure 8 (middle)
  p3,  # Figure 8 (right)
  ncol = 3
)
\end{example}

>>>>

\abstract{
Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes linearity and normality). Unfortunately, it can be challenging to understand the results of such models and explain them to management. Partial dependence plots offer a simple solution. Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be
more easily understood. These plots are especially useful in explaining the output from black box models. In this paper, we introduce \pkg{pdp}, a general R package for constructing partial dependence plots.
}

@@

\strong{Note:} as mentioned above, \pkg{pdp} relies on \pkg{lattice} for its graphics. \pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}. \pkg{grid} graphics behave a little differently than traditional R graphics, and two points are worth making (see \code{?lattice} for more details):
\begin{enumerate}
  \item \pkg{lattice} functions return a \code{"trellis"} object, but do not display it; the \code{print} method produces the actual display. However, due to R's automatic printing rule, the result is automatically printed when using these functions in the command line. If \code{plotPartial} is called inside of \code{source} or inside a loop (e.g., \code{for} or \code{while}), an explicit \code{print} statement is required to display the resulting graph.
  \item Setting graphical parameters via \code{par} typically has no effect on \pkg{lattice} plots. Instead, \pkg{lattice} provides its own \code{trellis.par.set} function for modifying graphical parameters.
\end{enumerate}
A consequence of the second point is that the \code{par} function cannot be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots. Simple solutions are available in packages \CRANpkg{latticeExtra} \citep{latticeExtra-pkg} and \CRANpkg{gridExtra} \citep{gridExtra-pkg}. For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{pdp} plots on a single page.

@@

When additional arguments are necessary, the user will be prompted with an informative error message. For instance, the user may occasionally see the following:
\begin{example}
Error: The training data could not be extracted from object. Please supply
the raw training data using the `train` argument in the call to `partial`.
\end{example}
This will occur, for example, when using partial with the \code{"xgb.Booster"} class (see Section~\ref{sec:xgboost}). In fact, using \code{partial} with any object that does not store the original training data will result in the above error message.

The \code{partial} function also supports object of class \code{"train"} produced using the \code{train} function from the well-known \CRANpkg{caret} package \citep{caret-pkg}. This means that \code{partial} can be used with any classification or regression model that has been fit using \pkg{caret}'s \code{train} function; see \url{http://topepo.github.io/caret/available-models.html} for a current list of models supported by \pkg{caret}. An example is given in Section~\ref{sec:xgboost}.

For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}. These data contain the median value of owner-occupied homes in 506 U.S. census tracts in the Boston area, along with 13 independent variables such as the per capita crime rate by town. We begin by loading the data and omitting unimportant columns.
\begin{example}
data(BostonHousing2, package = "mlbench")  # mlbench must be installed!
boston <- BostonHousing2[, -c(1, 2, 5)]
\end{example}
Next, we fit a random forest to the entire data set with default tuning parameters and 500 trees:
\begin{example}
library(randomForest)
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # Figure 1
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}). The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk uses the random forest model to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}. The \code{grid.arrange} function is then used to display three PDPs, which make use of various \code{plotPartial} options\footnote{See Section~\ref{sec:computational} for an example of how to add a label to the colorkey in these types of graphs.}, on the same graph. The result is displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
rwb <- colorRampPalette(c("red", "white", "blue"))
pd <- partial(boston.rf, pred.var = c("lstat", "rm"))
grid.arrange(
  # Figure 3 (left)
  plotPartial(pd),
  # Figure 3 (middle)
  plotPartial(pd, contour = TRUE, col.regions = rwb),
  # Figure 3 (right)
  plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
              colorkey = TRUE, screen = list(z = -20, x = -60)),
  ncol = 3
)
\end{example}
\strong{Note:} the default color map for level plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Addressing computational concerns
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Addressing computational concerns}
\label{sec:computational}

Constructing PDPs can be quite computationally expensive. Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option, then uses \code{grid.arrange} to display all three PDPs on the same page; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
grid.arrange(
  # Figure 5 (left)
  partial(boston.rf, "rm", plot = TRUE),
  # Figure 5 (middle)
  partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),
  # Figure 5 (right)
  partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),
  ncol = 3
)
\end{example}

@@

Setting up a parallel backend is rather straightforward. To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on both Windows and Unix-like systems using the \pkg{doParallel} package.
\begin{example}
library(doParallel)  # load the parallel backend
cl <- makeCluster(4)  # use 4 workers
registerDoParallel(cl)  # register the parallel backend
\end{example}
Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} and \code{paropts} options and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded\footnote{Notice we have to pass the names of external packages that the tasks depend on via the \code{paropts} argument; in this case, \code{"earth"}. See \code{?plyr::adply} for details.}. This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel. The last three lines of code simply add a label to the colorkey. The result is displayed in Figure~\ref{fig:partial_par}. \strong{Note:} it is considered good practice to shut down the workers by calling \code{stopCluster} when finished.
\begin{example}
partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
        chull = TRUE, parallel = TRUE, paropts = list(.packages = "earth"))  # Figure 6
stopCluster(cl)  # good practice

# Add a label to the colorkey
lattice::trellis.focus("legend", side = "right", clipp.off = TRUE, highlight = FALSE)
grid::grid.text("ozone", x = 0.2, y = 1.05, hjust = 0.5, vjust = 1)
lattice::trellis.unfocus()
\end{example}

@@

To round out our discussion, we provide one last example using a recently popular (and successful!) machine learning tool. XGBoost, short for eXtreme Gradient Boosting, is a popular library providing optimized distributed gradient boosting that is specifically designed to be highly efficient, flexible and portable. The associated R package \pkg{xgboost} has been used to win a number of \href{https://www.kaggle.com/}{Kaggle competitions}. It has been shown to be many times faster than the well-known \pkg{gbm} package. However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for constructing PDPs. Fortunately, the \pkg{pdp} package can be used to fill this gap.

For illustration, we return to the Boston housing data. The code chunk below uses \pkg{caret} to train an XGBoost model to the \code{boston} data frame from Section~\ref{sec:boston} over a grid of tuning values (see \code{?xgboost::xgboost} for an explanation of the various tuning parameters). The optimal combination of tuning parameter values is selected via 5-fold cross-validation.
\begin{example}
library(caret)

@@

# Grid of tuning values for XGBoost.
xgb.grid <- expand.grid(nrounds = c(100, 500, 1000, 2000, 5000),
                        max_depth = 1:6,
                        eta = c(0.001, 0.01, 0.1, 0.5, 1),
                        gamma = 0,
                        colsample_bytree = 1,
                        min_child_weight = 1)

# Train an XGBoost model
set.seed(202)
boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
                    y = boston$cmedv,
                    method = "xgbTree",
                    metric = "RMSE",
                    trControl = ctrl,
                    tuneGrid = xgb.grid)
\end{example}
The model with the smallest root-mean-squared error (RMSE) had \code{nrounds = 2000}, \code{max\_depth = 3}, \code{eta = 0.01}, and achieved a cross-validated RMSE and $R^2$ of 2.819 and 0.905, respectively. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display all three PDPs on the same page. The results are displayed in Figure~\ref{fig:boston_xgb}.
\begin{example}
grid.arrange(
  # Figure 8 (left)
  partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE),
  # Figure 8 (middle)
  partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE),
  # Figure 8 (right)
  partial(boston.xgb, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE),
  ncol = 3
)
\end{example}
